---
title: "Practicum 2 part 1"
output: html_notebook
---

Team members:
Aditya Kekare (kekare.a@northeastern.edu) - 002135512
Rushikesh Mokashi (mokashi.r@northeastern.edu) - 001098615

1.3 ERD Schema
```{r}
img_url <- "https://i.imgur.com/G6K9qqF.png"
knitr::include_graphics(img_url)
```
```{r}
library(RSQLite)
dbfile = "part1.sqlite"

dbcon <- dbConnect(RSQLite::SQLite(), dbfile)

library(XML)
fn <- "pubmed_sample.xml"
xmlObj <- xmlParse(fn)
```

```{sql connection=dbcon}
PRAGMA foreign_keys = ON
```

1.4 Creating Tables 
```{sql connection=dbcon}
DROP TABLE IF EXISTS Article;
```

```{sql connection=dbcon}
DROP TABLE IF EXISTS Journal;
```

```{sql connection=dbcon}
DROP TABLE IF EXISTS Author;
```

```{sql connection=dbcon}
DROP TABLE IF EXISTS AuthorList;
```

```{sql connection=dbcon}
CREATE TABLE IF NOT EXISTS Article(
pmid VARCHAR(20) PRIMARY KEY, 
Title VARCHAR(255),
dateCreated TEXT,
issn VARCHAR(255)
);
```

```{sql connection=dbcon}
CREATE TABLE IF NOT EXISTS Journal(
id INT PRIMARY KEY,
issn VARCHAR(255),
volume INT,
issue INT,
title VARCHAR(255),
pubDate TEXT);
```

```{sql connection=dbcon}
CREATE TABLE IF NOT EXISTS Author(
id INT PRIMARY KEY,
firstName VARCHAR(255),
lastName VARCHAR(255),
initials VARCHAR(5),
affiliation VARCHAR(255));
```

```{sql connection=dbcon}
CREATE TABLE IF NOT EXISTS AuthorList(
id INT PRIMARY KEY,
authorId INT,
pmid VARCHAR(20),
FOREIGN KEY(authorId) REFERENCES Author(id),
FOREIGN KEY(articleId) REFERENCES Article(pmid));
```

1.5 
Loading data from xml into dataframes and then populating tables


Article Dataframe fill from XML

To handle the date, we have chosen to represent it as %Y-%M-$D format for DateCreated. We have chosen to do so, as it
makes processing dates easier in Part 2 of data warehousing.
```{r}
article_df <- data.frame(PMID=character(),
                 Title=character(),
                 dateCreated=character(),
                 issn=character(),
                 stringsAsFactors=FALSE)
articles <- xpathSApply(xmlObj, "//PubmedArticle")

for(i in 1:length(articles)) {
  article_i <- xmlParseString(toString.XMLNode(articles[[i]]))
  
  pmid_i <- xpathSApply(article_i, "//MedlineCitation/PMID")
  title_i <- xpathSApply(article_i, "//MedlineCitation/Article/ArticleTitle")
  issn_i <- xpathSApply(article_i, "//MedlineCitation/Article/Journal/ISSN")
  
  xml_date_i <- xmlParseString(toString.XMLNode(xpathSApply(article_i, "//MedlineCitation/DateCreated")))
  year_i <- xpathSApply(xml_date_i, "//Year")
  month_i <- xpathSApply(xml_date_i, "//Month")
  day_i <- xpathSApply(xml_date_i, "//Day")
  
  date_i <- paste(xmlValue(year_i), xmlValue(month_i), xmlValue(day_i), sep="-")
  
  article_df[nrow(article_df) + 1,] = c(xmlValue(pmid_i),xmlValue(title_i), date_i, xmlValue(issn_i))
}
```

Journal Dataframe fill from XML

To handle PubDate we chose just to keep the year, as we observed that all dates have atleast the year field present.
Month and day were missing from some entries in the XML, so we decided to not keep those. 
```{r}
journal_df <- data.frame(issn=character(),
                 volume=integer(),
                 issue=character(),
                 title=character(),
                 pubDate=character(),
                 stringsAsFactors=FALSE)

articles <- xpathSApply(xmlObj, "//PubmedArticle")

for(i in 1:length(articles)) {
  article_i <- xmlParseString(toString.XMLNode(articles[[i]]))
  
  issn_i <- xpathSApply(article_i, "//Article/Journal/ISSN")
  title_i <- xpathSApply(article_i, "//Article/Journal/Title")
  volume_i <- xpathSApply(article_i, "//Article/Journal/JournalIssue/Volume")
  issue_i <- xpathSApply(article_i, "//Article/Journal/JournalIssue/Issue")

  xml_date_i <- xmlParseString(toString.XMLNode(xpathSApply(article_i, "//Article/Journal/JournalIssue/PubDate")))
  year_i <- xmlValue(xpathSApply(xml_date_i, "//Year"))
  if(length(year_i) == 0) {
    xml_date_i <- xmlValue(xpathSApply(article_i, "//Article/Journal/JournalIssue/PubDate/MedlineDate"))
    year_i <- substring(xml_date_i, 1, 4)
  }

  journal_df[nrow(journal_df) + 1,] = c( xmlValue(issn_i),xmlValue(volume_i), xmlValue(issue_i),  xmlValue(title_i), year_i)
}
journal_df <- unique(journal_df)
journal_df$id <- 1:nrow(journal_df)
```

Author dataframe Fill from XML
```{r}
author_df <- data.frame(firstName=character(),
                 lastName=character(),
                 initials=character(),
                 affiliation=character())

articles <- xpathSApply(xmlObj, "//PubmedArticle")

for(i in 1:length(articles)) {
  article_i <- xmlParseString(toString.XMLNode(articles[[i]]))
  authors <- xpathSApply(article_i, "//AuthorList/Author")
  
  for(j in 1: length(authors)){
    author_j <- xmlParseString(toString.XMLNode(authors[[j]]))
    firstName_j <- xmlValue(xpathSApply(author_j, "//Author/ForeName"))
    lastName_j <- xmlValue(xpathSApply(author_j, "//Author/LastName"))
    initials_j <- xmlValue(xpathSApply(author_j, "//Author/Initials"))
    affiliation_j <- xmlValue(xpathSApply(author_j, "//Author/Affiliation"))
    if(is.list(affiliation_j)){
      affiliation_j <- "NA"
    }
    author_df[nrow(author_df) + 1,] = c(firstName_j, lastName_j, initials_j, affiliation_j)
  }

}
author_df <- author_df[!duplicated(author_df[,c('firstName', 'lastName', 'initials')]),]
author_df$id <- 1:nrow(author_df)
author_df$id <- as.numeric(author_df$id)
```

AuthorList dataframe fill from XML
```{r}
authorList_df <- data.frame(pmid=character(),
                 authorId=integer())

articles <- xpathSApply(xmlObj, "//PubmedArticle")

for(i in 1:length(articles)) {
  article_i <- xmlParseString(toString.XMLNode(articles[[i]]))
  authors_i <- xpathSApply(article_i, "//AuthorList/Author")
  pmid_i <- xmlValue(xpathSApply(article_i, "//MedlineCitation/PMID"))
  
  for(j in 1:length(authors_i)) {
    author_j <- xmlParseString(toString.XMLNode(authors_i[[j]]))
    firstName_j <- xmlValue(xpathSApply(author_j, "//Author/ForeName"))
    lastName_j <- xmlValue(xpathSApply(author_j, "//Author/LastName"))
    initials_j <- xmlValue(xpathSApply(author_j, "//Author/Initials"))
    
    author_id <- NA
    for(i in 1:nrow(author_df)) {       # for-loop over rows
      if(firstName_j == author_df[i,1] && lastName_j == author_df[i,2] && initials_j == author_df[i,3]) {
        author_id <- author_df[i,5]
        break
      }
    }
    authorList_df[nrow(authorList_df) + 1,] = c(pmid_i, author_id)
  }
}
authorList_df$authorId <- as.numeric(authorList_df$authorId)
```

Checking Dataframes
```{r}
print(article_df)
print(journal_df)
print(author_df)
print(authorList_df)
```

Writing Dataframes to the database
```{r}
dbWriteTable(dbcon, name="journal", value = journal_df, append = TRUE, row.names = FALSE)
dbWriteTable(dbcon, name="author", value = author_df, append = TRUE, row.names = FALSE)
dbWriteTable(dbcon, name="article", value = article_df, append = TRUE, row.names = FALSE)
dbWriteTable(dbcon, name="authorlist", value = authorList_df, append = TRUE, row.names = FALSE)
```

Checking data inside tables 

```{sql connection=dbcon}
SELECT * FROM AuthorList;
```

```{sql connection=dbcon}
SELECT * FROM Journal;
```

```{sql connection=dbcon}
SELECT * FROM Author;
```

```{sql connection=dbcon}
SELECT * FROM Article;
```

