---
title: "Practicum 2 part 2,3"
output: html_notebook
---

Team members:
Aditya Kekare (kekare.a@northeastern.edu) - 002135512
Rushikesh Mokashi (mokashi.r@northeastern.edu) - 001098615

```{r}
library(RMySQL)

db_user <- 'root'
db_password <- 'root'
db_name <- 'Practicum2'
db_host <- 'localhost' 
db_port <- 3306
dbcon_mysql <-  dbConnect(MySQL(), user = db_user, password = db_password,
                 dbname = db_name, host = db_host, port = db_port)

library(RSQLite)
dbfile = "part1.sqlite"

dbcon_sqlite <- dbConnect(RSQLite::SQLite(), dbfile)

library(XML)
fn <- "pubmed_sample.xml"
xmlObj <- xmlParse(fn)
```

```{sql connection=dbcon_mysql}
CREATE DATABASE IF NOT EXISTS Practicum2;
```

```{sql connection=dbcon_mysql}
USE Practicum2;
```



2 Creating Dimension and Fact tables
```{sql connection=dbcon_mysql}
DROP TABLE IF EXISTS AuthorFact;
```

```{sql connection=dbcon_mysql}
DROP TABLE IF EXISTS ArticleDimension;
```

```{sql connection=dbcon_mysql}
CREATE TABLE IF NOT EXISTS ArticleDimension (
pmid VARCHAR(20) PRIMARY KEY,
title VARCHAR(225),
date_created VARCHAR(30),
issn VARCHAR(255)
);
```

```{sql connection=dbcon_mysql}
CREATE TABLE IF NOT EXISTS AuthorFact ( 
id INT, 
articleid VARCHAR(20),
author_name VARCHAR(255),
num_articles INT,
avg_articles_year DOUBLE,
PRIMARY KEY(id, articleid),
FOREIGN KEY(articleid) REFERENCES ArticleDimension(pmid)
);
```



Loading data to be added to AuthorFact inside author_df dataframe
```{r}
query1 <- "
SELECT 
  AuthorList.authorId,
  AuthorList.pmid as articleid,
  (Author.firstname || ' ' || Author.lastname) AS author_name
FROM AuthorList
LEFT JOIN Author ON AuthorList.authorId = Author.id
"

#  Article.pmid as articleid,
#   (firstname || ' ' || lastname) AS author_name,
query <- "
SELECT 
  Author.id as authorId,
  COUNT(Article.pmid) AS num_articles,
  ROUND(COUNT(Article.pmid) * 1.0 / (
    SELECT COUNT(*)
    FROM
      (SELECT DISTINCT (SUBSTR(Article.dateCreated, 1, 4)) FROM Article)
  ),2) AS avg_articles_year
FROM Author
INNER JOIN AuthorList ON Author.id = AuthorList.authorId
INNER JOIN Article ON AuthorList.pmid = Article.pmid
GROUP BY Author.id
"
first_df <- dbGetQuery(dbcon_sqlite, query1)
author_df <- dbGetQuery(dbcon_sqlite, query)
final_df <- merge(first_df, author_df, by=c("authorId"))
print(first_df)
print(author_df)
colnames(final_df)[1] <- "id"
print(final_df)
```

Loading data to be added to ArticleDimension inside article_df dataframe
```{r}
query <- "
SELECT pmid,
  Title as title,
  dateCreated as date_created,
  issn
FROM 
  Article;
"
article_df <- dbGetQuery(dbcon_sqlite, query)
print(article_df)
```

Writing the dataframes into the database.
```{r}
dbExecute(dbcon_mysql, "SET GLOBAL local_infile = TRUE;")
dbWriteTable(dbcon_mysql, name="articledimension", value = article_df, append = TRUE, row.names = FALSE)
dbWriteTable(dbcon_mysql, name="authorfact", value = final_df, append = TRUE, row.names = FALSE)
```


Checking the contents of the tables.
```{sql connection=dbcon_mysql}
SELECT * FROM AuthorFact;
```

```{sql connection=dbcon_mysql}
SELECT * FROM ArticleDimension;
```

Query to check top 10 authors according to the number of articles.
```{sql connection=dbcon_mysql}
SELECT DISTINCT id,author_name AS author, num_articles AS 'Number of Articles'
FROM AuthorFact
ORDER BY num_articles DESC
LIMIT 10;
```
